\documentclass[11pt]{article}
\usepackage[margin=1in]{geometry}
\usepackage{times}
\usepackage{hyperref}
\usepackage{xcolor}

\hypersetup{
  colorlinks=true,
  linkcolor=blue,
  urlcolor=blue
}

\begin{document}

\noindent
\textbf{\large Response to Technical Check --- Manuscript ID: afe5f6bf-7c9a-47d8-bddd-c3fadbaae817 v1.0}

\medskip
\noindent
\textbf{Manuscript Title:} Perceive, Plan, Act, Self-Correct: An Architectural Framework for Goal-Directed Agentic AI Systems

\medskip
\noindent
\textbf{Journal:} Artificial Intelligence Review (Springer Nature)

\medskip
\noindent
\textbf{Author:} Hameem Mahdi

\medskip
\noindent
\textbf{Date:} 10 April 2026

\bigskip
\noindent
Dear Editorial Office,

\medskip
\noindent
Thank you for completing the technical check of our manuscript. Below we address the requested amendment.

\bigskip
\noindent
\rule{\textwidth}{0.4pt}

\medskip
\noindent
\textbf{Point 1:} \emph{Please provide a Funding Declaration in the manuscript. If there was no Funding, please state so.}

\medskip
\noindent
\textbf{Response:} A Funding subsection has been added to the Declarations section of the manuscript. The statement reads:

\medskip
\begin{quote}
\textit{``This research received no external funding.''}
\end{quote}

\noindent
This appears in Section~\textbf{Declarations~$\rightarrow$~Funding} of the revised manuscript.

\medskip
\noindent
\rule{\textwidth}{0.4pt}

\bigskip
\noindent
No other changes were made to the manuscript. We hope this amendment satisfactorily addresses the technical check requirement.

\bigskip
\noindent
Best regards,\\
Hameem Mahdi, B.S.C.S., M.S.E., Ph.D

\end{document}
